---
title: "Analysis of Real Estate Market Trends  "
author: "Aden Koziol"
date: "3/31/2025"
output: html_document
---

## Description of the Data Set  
The dataset `realtor-data.csv` contains information about real estate listings, including variables such as price, number of bedrooms/bathrooms, square footage, location, and other relevant features.  

## Data Cleaning Steps  

```{r data-loading}
library(tidyverse)
library(knitr)

realtor_data <- read_csv("realtor-data.csv")

head(realtor_data)

# Missing values
missing_values <- colSums(is.na(realtor_data))
kable(missing_values, caption = "Missing Values per Column")

# Remove rows with missing critical data (e.g., price, bedrooms)
cleaned_data <- realtor_data %>%
  drop_na(price, bed, bath, house_size)

# Remove duplicates
cleaned_data <- distinct(cleaned_data)

# Convert price to numeric (if stored as character)
cleaned_data <- cleaned_data %>%
  mutate(price = as.numeric(gsub("[\\$,]", "", price)))

# Filter out unrealistic prices (e.g., below $10,000 or above $10M)
cleaned_data <- cleaned_data %>%
  filter(price >= 10000 & price <= 10000000)
```

## Marginal Summaries
```{r}
# Summary statistics for numeric variables
summary_stats <- cleaned_data %>%
  select(price, bed, bath, house_size) %>%
  summary()

kable(summary_stats, caption = "Summary Statistics")

# Distribution of categorical variables (e.g., state, city)
if ("state" %in% colnames(cleaned_data)) {
  state_counts <- cleaned_data %>%
    count(state) %>%
    arrange(desc(n))
  
  kable(state_counts, caption = "Property Count by State")
}
```

## Questions To Be Addressed

- How does property price vary by things like state and city?

- What is the relationship between square feet and price?

- How does number of bedrooms/bathrooms affect price?

- Can we predict property prices based on things like square feet, bedrooms, and/or location?